---
title: "Online Retail Customer Analytics"
author: "Madhu Mitha"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

# Project Overview

This project presents an end-to-end analysis of online retail transaction data to understand customer purchasing behavior, revenue performance, geographic activity, product performance, cancellations, data quality, and transaction patterns.

The analysis was completed in R using data wrangling, exploratory data analysis, aggregation, feature engineering, and visualization techniques to transform raw transaction data into business-focused insights.

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(dplyr)
library(ggplot2)
```

# Data Import

The dataset contains transaction-level records from a UK-based online retailer, including invoice information, products, quantities, prices, customer identifiers, and customer countries.

```{r data-import}
retail <- read.csv("../data/Online_Retail.csv")

dim(retail)
head(retail)
str(retail)
summary(retail)
```

# Data Preparation

A transaction-level revenue measure was created by multiplying the quantity purchased by the unit price.

```{r transaction-value}
retail <- retail %>%
  mutate(
    TransactionValue = Quantity * UnitPrice
  )

summary(retail$TransactionValue)
```

# Date & Time Feature Engineering

The invoice timestamp was converted into a date-time object and separated into additional analytical features.

```{r date-features}
invoice_datetime <- strptime(
  retail$InvoiceDate,
  format = "%m/%d/%Y %H:%M",
  tz = "GMT"
)

retail <- retail %>%
  mutate(
    InvoiceDateParsed = as.Date(invoice_datetime),
    InvoiceDayOfWeek = weekdays(InvoiceDateParsed),
    InvoiceHour = as.numeric(format(invoice_datetime, "%H")),
    InvoiceMonth = as.numeric(format(invoice_datetime, "%m"))
  )
```

These features make it possible to analyze customer activity across dates, weekdays, hours, and months.

# Data Quality Analysis

Missing values were evaluated across all variables to identify potential data-quality issues.

```{r missing-values}
missing_percentage <- colMeans(is.na(retail)) * 100

missing_percentage
```

Missing Customer ID records were also analyzed by country.

```{r missing-customer-country}
missing_customer_by_country <- retail %>%
  filter(is.na(CustomerID)) %>%
  count(Country, sort = TRUE)

missing_customer_by_country
```

This analysis helps identify where incomplete customer information is concentrated and where additional validation may be required before customer-level analysis.

# Geographic Analysis

## Transaction Activity by Country

Transaction activity was aggregated by country to understand the geographic distribution of the retailer's business.

```{r country-transactions}
country_transactions <- retail %>%
  count(Country, name = "Transactions") %>%
  mutate(
    Percentage = 100 * Transactions / nrow(retail)
  ) %>%
  arrange(desc(Transactions))

country_transactions %>%
  filter(Percentage > 1)
```

## Transaction Value by Country

Total transaction value was calculated for each country to identify markets generating significant revenue.

```{r country-revenue}
country_revenue <- retail %>%
  group_by(Country) %>%
  summarise(
    TotalTransactionValue =
      sum(TransactionValue, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(TotalTransactionValue))

country_revenue %>%
  filter(TotalTransactionValue > 130000)
```

# Geographic Visualization

```{r geographic-visualization}
top_countries <- country_revenue %>%
  slice_max(
    order_by = TotalTransactionValue,
    n = 10
  )

ggplot(
  top_countries,
  aes(
    x = reorder(Country, TotalTransactionValue),
    y = TotalTransactionValue
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top Countries by Transaction Value",
    x = "Country",
    y = "Total Transaction Value"
  ) +
  theme_minimal()
```

# Time-Based Analysis

## Transaction Activity by Day of Week

```{r weekday-activity}
weekday_transactions <- retail %>%
  group_by(InvoiceDayOfWeek) %>%
  summarise(
    Transactions = n(),
    .groups = "drop"
  ) %>%
  mutate(
    TransactionPercentage =
      100 * Transactions / sum(Transactions)
  )

weekday_transactions
```

## Transaction Value by Day of Week

```{r weekday-value}
weekday_value <- retail %>%
  group_by(InvoiceDayOfWeek) %>%
  summarise(
    TransactionValue =
      sum(TransactionValue, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    ValuePercentage =
      100 * TransactionValue /
      sum(TransactionValue)
  )

weekday_value
```

## Monthly Transaction Value

```{r monthly-value}
monthly_value <- retail %>%
  group_by(InvoiceMonth) %>%
  summarise(
    TransactionValue =
      sum(TransactionValue, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    ValuePercentage =
      100 * TransactionValue /
      sum(TransactionValue)
  )

monthly_value
```

## Monthly Revenue Visualization

```{r monthly-revenue-chart}
ggplot(
  monthly_value,
  aes(
    x = InvoiceMonth,
    y = TransactionValue
  )
) +
  geom_col() +
  scale_x_continuous(breaks = 1:12) +
  labs(
    title = "Monthly Transaction Value",
    x = "Month",
    y = "Transaction Value"
  ) +
  theme_minimal()
```

# Australia Transaction Analysis

The date with the highest transaction activity for Australian customers was identified.

```{r australia-analysis}
australia_transactions <- retail %>%
  filter(Country == "Australia") %>%
  count(
    InvoiceDateParsed,
    name = "Transactions"
  ) %>%
  arrange(desc(Transactions))

head(australia_transactions, 1)
```

# Website Activity Analysis

Hourly transaction activity was evaluated to understand customer usage patterns during the IT team's operating window.

```{r hourly-analysis}
hourly_activity <- retail %>%
  filter(
    InvoiceHour >= 7,
    InvoiceHour <= 20
  ) %>%
  group_by(InvoiceHour) %>%
  summarise(
    Transactions = n(),
    .groups = "drop"
  ) %>%
  arrange(InvoiceHour)

hourly_activity
```

Two consecutive hours were evaluated to identify the lowest-activity maintenance window.

```{r maintenance-window}
maintenance_windows <- data.frame(
  StartHour = hourly_activity$InvoiceHour[-nrow(hourly_activity)],
  EndHour = hourly_activity$InvoiceHour[-1],
  TwoHourTransactions =
    hourly_activity$Transactions[-nrow(hourly_activity)] +
    hourly_activity$Transactions[-1]
)

maintenance_windows %>%
  arrange(TwoHourTransactions) %>%
  head(1)
```

## Hourly Activity Visualization

```{r hourly-chart}
ggplot(
  hourly_activity,
  aes(
    x = InvoiceHour,
    y = Transactions
  )
) +
  geom_line() +
  geom_point() +
  labs(
    title = "Transaction Activity by Hour",
    x = "Hour of Day",
    y = "Number of Transaction Records"
  ) +
  theme_minimal()
```

# Customer Analysis

Customer-level analysis excludes records where Customer ID is unavailable.

```{r customer-data}
customer_data <- retail %>%
  filter(!is.na(CustomerID))
```

## Most Active Customer

Customers were ranked according to the number of transaction records associated with their Customer ID.

```{r active-customer}
customer_activity <- customer_data %>%
  count(
    CustomerID,
    name = "TransactionRecords"
  ) %>%
  arrange(desc(TransactionRecords))

head(customer_activity, 10)
```

## Highest-Value Customer

Customers were also ranked by total transaction value.

```{r valuable-customer}
customer_value <- customer_data %>%
  group_by(CustomerID) %>%
  summarise(
    TotalTransactionValue =
      sum(TransactionValue, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(TotalTransactionValue))

head(customer_value, 10)
```

## Unique Customers

```{r unique-customers}
unique_customer_count <-
  n_distinct(customer_data$CustomerID)

unique_customer_count
```

# Repeat Purchase Analysis

Customer purchase dates were analyzed to estimate the number of days between consecutive shopping visits.

Multiple product rows can belong to the same customer purchase, so unique customer-date combinations are used before calculating the time between visits.

```{r repeat-purchase-analysis}
customer_purchase_dates <- customer_data %>%
  filter(!is.na(InvoiceDateParsed)) %>%
  distinct(
    CustomerID,
    InvoiceDateParsed
  ) %>%
  arrange(
    CustomerID,
    InvoiceDateParsed
  )

repeat_purchase_intervals <-
  customer_purchase_dates %>%
  group_by(CustomerID) %>%
  mutate(
    DaysBetweenPurchases =
      as.numeric(
        InvoiceDateParsed -
        lag(InvoiceDateParsed)
      )
  ) %>%
  ungroup() %>%
  filter(
    !is.na(DaysBetweenPurchases),
    DaysBetweenPurchases > 0
  )

average_days_between_purchases <-
  mean(
    repeat_purchase_intervals$DaysBetweenPurchases,
    na.rm = TRUE
  )

average_days_between_purchases
```

# Returns & Cancellation Analysis

Negative quantity values are used to identify cancelled transaction records.

The cancellation rate for French customers was calculated as the number of cancelled transaction records divided by the total number of French transaction records.

```{r france-cancellation}
france_transactions <- retail %>%
  filter(Country == "France")

france_cancellation_rate <-
  100 *
  sum(
    france_transactions$Quantity < 0,
    na.rm = TRUE
  ) /
  nrow(france_transactions)

france_cancellation_rate
```

# Product Performance Analysis

Products were ranked according to total transaction value to identify the retailer's highest revenue-generating items.

```{r product-performance}
product_revenue <- retail %>%
  group_by(
    StockCode,
    Description
  ) %>%
  summarise(
    TotalTransactionValue =
      sum(TransactionValue, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(TotalTransactionValue))

head(product_revenue, 10)
```

## Top Product Visualization

```{r product-chart}
top_products <- product_revenue %>%
  filter(!is.na(Description)) %>%
  slice_max(
    order_by = TotalTransactionValue,
    n = 10
  )

ggplot(
  top_products,
  aes(
    x = reorder(
      Description,
      TotalTransactionValue
    ),
    y = TotalTransactionValue
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top Products by Transaction Value",
    x = "Product",
    y = "Total Transaction Value"
  ) +
  theme_minimal()
```

# Germany Transaction Value Distribution

The distribution of transaction values for German customers was evaluated to understand the spread of transaction-level values.

```{r germany-analysis}
germany_transactions <- retail %>%
  filter(Country == "Germany")

hist(
  germany_transactions$TransactionValue,
  main = "Distribution of Transaction Values - Germany",
  xlab = "Transaction Value"
)
```

# Key Analytical Outputs

The completed analysis provides insight into:

- Geographic concentration of transaction activity
- Revenue contribution across countries
- Customer activity and customer value
- Repeat-purchase behavior
- Product-level revenue performance
- Cancellation behavior
- Missing customer information
- Weekday and monthly transaction patterns
- Hourly customer activity
- Potential low-activity maintenance periods

# Skills Demonstrated

- R
- dplyr
- ggplot2
- Data Wrangling
- Exploratory Data Analysis
- Feature Engineering
- Descriptive Statistics
- Customer Analytics
- Revenue Analysis
- Product Analytics
- Data Quality Analysis
- Data Visualization

# Conclusion

This project demonstrates an end-to-end retail analytics workflow, beginning with raw transaction data and progressing through data preparation, quality assessment, feature engineering, customer analysis, geographic analysis, revenue analysis, product analysis, and visualization.

The analysis shows how transaction-level retail data can be transformed into actionable business insights that support customer understanding, revenue monitoring, merchandising decisions, and operational planning.
